---
title: "African Languages: A Statistical Overview"
author: "Abdallah Ramadhan"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(patchwork)
```

```{r define-theme, include=FALSE}
theme_african <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 13, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "grey40"),
    plot.caption  = element_text(size = 8, hjust = 1, color = "grey50"),
    axis.title    = element_text(size = 10),
    axis.text     = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.margin   = margin(15, 15, 15, 15)
  )

africa_colours <- c(
  "Niger–Congo"              = "#E07B39",
  "Nilo-Saharan"             = "#4A90A4",
  "Afroasiatic"              = "#C0392B",
  "Indo-European"            = "#7D3C98",
  "Khoe–Kwadi"               = "#27AE60",
  "Kxʼa"                     = "#2ECC71",
  "Ubangian"                 = "#F39C12",
  "Tuu"                      = "#1ABC9C",
  "Austronesian"             = "#3498DB",
  "Kongo-Creole/Pidgin"      = "#E74C3C",
  "Arabic-Creole/Pidgin"     = "#D4AC0D",
  "English-Creole/Pidgin"    = "#85929E",
  "French-Creole/Pidgin"     = "#5D6D7E",
  "Portuguese-Creole/Pidgin" = "#A04000"
)
```

## 1. Introduction

Africa is the world's most linguistically diverse continent, home to an
estimated 2,000 languages spanning dozens of distinct language families.
This report analyses a dataset of 502 African languages across 51 countries,
exploring three core questions:

- Which country in Africa has the largest number of spoken languages?
- Which language family has the highest density of speakers?
- Are there languages that cut across multiple countries?

---

## 2. Data Preparation

The dataset was loaded from a CSV file containing 796 rows and 4 variables:
`language`, `family`, `native_speakers`, and `country`.

### 2.1 Loading the Data

```{r load-data}
data <- read_csv("../data/africa_languages.csv",
                 locale = locale(encoding = "UTF-8"))
glimpse(data)
```

### 2.2 Data Quality Issues Identified

Before analysis, the following issues were identified and resolved:

| Issue | Action Taken |
|---|---|
| Mojibake encoding errors in language names | Fixed by loading with UTF-8 encoding |
| `Afro-Asiatic` spelling inconsistency | Standardised to `Afroasiatic` |
| `Shabo` misclassified as `Language` family | Reclassified to `Nilo-Saharan` |
| `Bozo` listed under `Mande` | Moved to `Niger-Congo` |
| 9 creole/pidgin languages mislabelled | Reclassified under respective Creole/Pidgin families |
| 36 duplicate rows (same language, same country) | Removed using `distinct()` |

### 2.3 Cleaning Code

```{r clean-data}
data_clean <- data %>%
  mutate(family = case_when(
    language == "Shabo"                  ~ "Nilo-Saharan",
    language == "Bozo"                   ~ "Niger\u2013Congo",
    language == "Kituba"                 ~ "Kongo-Creole/Pidgin",
    language == "Nubi"                   ~ "Arabic-Creole/Pidgin",
    language == "Krio"                   ~ "English-Creole/Pidgin",
    language == "Pichinglis"             ~ "English-Creole/Pidgin",
    language == "Mauritian Creole"       ~ "French-Creole/Pidgin",
    language == "Seychellois Creole"     ~ "French-Creole/Pidgin",
    language == "Cape Verdean Creole"    ~ "Portuguese-Creole/Pidgin",
    family   == "Afro-Asiatic"           ~ "Afroasiatic",
    TRUE ~ family
  )) %>%
  distinct(language, country, .keep_all = TRUE)
```

### 2.4 Final Dataset Summary

```{r summary}
data_clean %>%
  summarise(
    total_rows       = n(),
    unique_languages = n_distinct(language),
    unique_countries = n_distinct(country),
    unique_families  = n_distinct(family)
  )
```

---

## 3. Exploratory Data Analysis

### 3.1 Distribution of Native Speakers

The `native_speakers` variable is heavily right-skewed, with a median of
150,000 but a mean of nearly 5 million — driven by a small number of
languages with massive speaker bases like Arabic and Swahili.

```{r distribution}
summary(data_clean$native_speakers)
```

A log transformation normalises the distribution and is used throughout
this analysis for visualisation and comparison purposes.

```{r log-histogram, fig.width=8, fig.height=4}
data_clean %>%
  mutate(log_speakers = log10(native_speakers)) %>%
  ggplot(aes(x = log_speakers)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Native Speakers (Log10 Scale)",
    x     = "Log10(Native Speakers)",
    y     = "Count"
  ) +
  theme_african
```

---

## 4. Research Findings

### 4.1 Which Country Has the Most Languages?

```{r chart1, fig.width=9, fig.height=6}
p1 <- data_clean %>%
  group_by(country) %>%
  summarise(num_languages = n_distinct(language)) %>%
  arrange(desc(num_languages)) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(country, num_languages), y = num_languages)) +
  geom_col(fill = "#E07B39", width = 0.7) +
  geom_text(aes(label = num_languages),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 110)) +
  labs(
    title    = "Top 10 African Countries by Number of Languages",
    subtitle = "Cameroon leads with extraordinary linguistic diversity",
    x        = NULL,
    y        = "Number of Languages",
    caption  = "Source: African Languages Dataset"
  ) +
  theme_african
p1
```

Cameroon leads with **95 languages**, followed by Congo (79) and
Nigeria (73). Cameroon's extraordinary diversity reflects its position
at the intersection of West and Central African linguistic zones,
earning it the nickname *"Africa in miniature"*. For organisations
working in education, health communication or community development,
this concentration of diversity means multilingual programming is not
optional — it is essential.

### 4.2 Which Language Family Has the Highest Speaker Density?

```{r family-table}
data_clean %>%
  group_by(family) %>%
  summarise(
    `Number of Languages` = n_distinct(language),
    `Median Speakers`     = scales::comma(median(native_speakers)),
    `Total Speakers`      = scales::comma(sum(native_speakers))
  ) %>%
  arrange(desc(`Number of Languages`)) %>%
  knitr::kable(caption = "Language Family Statistics")
```

```{r chart3, fig.width=9, fig.height=7}
p3 <- data_clean %>%
  ggplot(aes(x = reorder(family, native_speakers, median),
             y = native_speakers,
             fill = family)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_manual(values = africa_colours) +
  coord_flip() +
  labs(
    title    = "Native Speaker Distribution by Language Family",
    subtitle = "Log scale — showing spread and outliers within each family",
    x        = NULL,
    y        = "Native Speakers (log scale)",
    caption  = "Source: African Languages Dataset"
  ) +
  theme_african
p3
```

The answer depends on the metric used. **Afroasiatic** has the highest
median speakers per language (~22 million), driven by Arabic.
**Niger-Congo** contains the most languages (384) but most are small
communities, with a median of only 135,000 speakers — revealing a
striking paradox between linguistic diversity and speaker density.

### 4.3 Which Languages Span Multiple Countries?

```{r chart2, fig.width=10, fig.height=7}
p2 <- data_clean %>%
  group_by(language) %>%
  summarise(
    num_countries = n_distinct(country),
    family        = first(family)
  ) %>%
  filter(num_countries > 1) %>%
  arrange(desc(num_countries)) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(language, num_countries),
             y = num_countries,
             fill = family)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = num_countries),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = africa_colours) +
  scale_y_continuous(limits = c(0, 14)) +
  labs(
    title    = "Most Widespread African Languages",
    subtitle = "Top 15 languages by number of countries spanned",
    x        = NULL,
    y        = "Number of Countries",
    fill     = "Language Family",
    caption  = "Source: African Languages Dataset"
  ) +
  theme_african +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
p2
```

Of 502 languages, **155 (31%) span more than one country**. Arabic is
the most widespread at 12 countries, followed by Fulani at 10.
Notably, **347 languages (69%) are endemic to a single country** —
confined to one national context with no cross-border community to
sustain them. These languages are the most vulnerable to extinction
through policy neglect and demographic change.

> **For NGO programming:** Endemic languages warrant priority attention
> in cultural preservation, mother-tongue education, and community
> documentation initiatives.

---

## 5. Conclusion

This analysis reveals three defining features of African linguistic
diversity:

1. **Concentration of diversity** — Cameroon, Congo and Nigeria together
account for a disproportionate share of the continent's languages.
Organisations operating in these countries face the most complex
multilingual environments on the continent.

2. **The Niger-Congo paradox** — The largest family by language count
contains mostly small, localised communities with limited institutional
support. Scale of diversity does not translate to security of survival.

3. **Endemic fragility** — 69% of languages are confined to a single
country, highlighting the vulnerability of Africa's linguistic heritage
and the urgency of preservation efforts.

### Recommendations for Further Analysis

- Incorporate **language endangerment status** data from UNESCO to
identify which endemic languages are most at risk
- Map languages geographically to identify border communities split
by colonial boundaries
- Cross-reference with **education policy data** to assess which
languages have formal recognition in school systems

### Limitations

- Speaker counts represent a single point in time and may be outdated
- The dataset may not be exhaustive — some estimates put Africa's
languages at over 2,000
- Creole/Pidgin classification required manual research and judgement calls

---

*Report generated using R `r getRversion()` and
tidyverse `r packageVersion("tidyverse")`*
